---
title: "Introduction to momo"
description: >
  Learn how to get started with the basics of momo.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to momo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      error = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width=8, fig.height=7)
```


The goal of the *mo*vement *mo*del *momo* is to offer a fast and flexible model
and R package that allows estimating habitat preferences and fine-scale movement
patterns based on tagging data. To get started, make sure to install the most
recent package version from GitHub.


## Installation

The current version of *momo* (v0.0.1) requires R $>= 4.0.0$ and can be
downloaded from github as follows.

::: .momo-release
```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("tokami/momo")
```
:::

The development version can be installed with:

::: .momo-devel
```{r, eval = FALSE}
remotes::install_github("tokami/momo", ref = "dev")
```
:::



## Basic usage

After successful installation, the package can be loaded into the R environment.

```{r, eval = TRUE, echo = FALSE, fig.alt = ""}
library(momo)
```

A data set containing all required information can quickly be created by the
function `sim.momo`:

```{r, eval = TRUE, echo = FALSE, fig.alt = ""}
# Simulate a data set with tagging data
sim <- sim.momo()
```

The main function to estimate movement paramters is `fit.momo`:

```{r, eval = TRUE, echo = FALSE, fig.alt = ""}
# Fit the movement model
fit <- fit.momo(sim)
```

The package includes a variety of functions that allows to inspect the results:

```{r, eval = TRUE, echo = FALSE, fig.alt = "Simulated data and model predictions"}
# Plot the results
plotmomo.compare(sim = sim, fit = fit, bg = "white")
```

This example illustrates the basic workflow: prepare data (here based on
simulation), fit the model, and access results. For real applications, momo
supports multiple types of tagging data, customizable environmental covariates,
and both estimation and prediction features.



## A more realistic scenario

In a more realistic scenario, the tagging data and environmental fields are not
simulated but derived from actual observations, requiring additional data
processing before fitting *momo*. The [tutorial vignette](tutorial.html)
provides a step-by-step guide to applying *momo* to simulated data that closely
mimics real-world observations of skipjack tuna in the Eastern Pacific Ocean. It
illustrates the expected data formats and demonstrates how *momo* functions can
be used to prepare and structure observational datasets for model fitting.
