---
title: "Introduction to momo"
description: >
  Learn how to get started with the basics of momo.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to momo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

The goal of momo is to offer a flexible and fast model that can estimate habitat
preferences and fine-scale movement patterns based on tagging data.


## *Installing and loading *momo**
The current version of *momo* (v0.0.1) requires R $>= 4.0.0$ and can be downloaded
from github as follows.

```{r, echo = TRUE, include = TRUE, eval = FALSE}
install.packages("remotes")
remotes::install_github("tokami/momo/momo")
```

The development version can be installed with:

```{r, eval=FALSE, echo=TRUE}
install.packages("remotes")
remotes::install_github("tokami/momo/momo", ref = "dev")
```

The package is loaded into the R environment with:

```{r, eval=TRUE, echo=TRUE}
library(momo)
```



## *Tagging data*

A data set containing all required information can quickly be created by the
function `sim.momo`:

```{r, eval = FALSE}
# Simulate a data set with tagging data
sim <- sim.momo()
```



## *Fitting momo*

The main function to estimate movement paramters is `fit.momo`:

```{r, eval = FALSE}
# Fit the movement model
fit <- fit.momo(sim)
```



## *Inspecting results*

The package includes a variety of functions that allows to inspect the results:

```{r, eval = FALSE}
# Plot the results
plotmomo.compare(sim = sim, fit = fit)
```
