---
title: "Fine-scale movement modeling based on tagging data"
author: "Tobias K. Mildenberger and Anders Nielsen"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
    fig_caption: yes
    number_sections: true
    df_print: paged
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
vignette: >
  %\VignetteIndexEntry{Fine-scale movement modeling based on tagging data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: momo.bib
---

```{r ReaddataLoadLibraries, message=FALSE, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      cache = FALSE,
                      warning = FALSE,
                      eval = TRUE,
                      error = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      include = TRUE,
                      collapse = TRUE,
                      comment = "#>",
                      fig.show = "hold",
                      fig.width=8, fig.height=7)
```

This tutorial illustrates the application of the *momo* package to model
fine-scale movement based on tagging data.

**Outline:**

1. Prepare the data
2. Fit the model
3. Results
4. Advanced settings
5. Summary
6. References


The package is loaded into the R environment with:

```{r, eval=TRUE, echo=TRUE}
library(momo)
```

This introductory tutorial uses a simulated data set included in the package:
"skjepo". We load it into the R environment with the command `data(skjepo)`.

```{r, eval=TRUE, echo=FALSE}
data(skjepo)
```

This command loads not only the main list object `skjepo` (of class `momo.sim`),
which contains all required simulated data and can be used right away to fit the
model, but also four individual data sets: `skjepo.grid`, `skjepo.env`,
`skjepo.ctags`, and `skjepo.atags`. These data sets illustrate the typical
structure of raw input data and demonstrate how the functions in the *momo*
package can be used to convert them into the format required by the model as
demonstrated in the next section.

## *Prepare the data*
The package provides two main data sets: one with information on releases and
recoveries of conventional tags, and another with track data from archival tags.
These data sets can be prepared for analysis using the built-in functions
`prep.ctags` and `prep.atags`. Both functions allow users to specify which
columns contain the release and recapture times and locations, and automatically
convert date fields to the required format. Additional features, such as a speed
filter to flag implausible movements, are also available. For full details on
how to use each function, refer to their respective help pages (e.g.,
`help("prep.ctags")`).

```{r, eval=TRUE, echo=FALSE}
## Load ctags
ctags <- prep.ctags(skjepo.ctags,
                    names = c("date_time","date_caught",
                              "rel_lon","recap_lon",
                              "rel_lat","recap_lat"),
                    origin = "1899-12-30",
                    speed.limit = 200)
plotmomo.ctags(ctags, plot.land = TRUE)

## Load atags
atags <- prep.atags(skjepo.atags,
                    names = c("time","mptlon","mptlat"),
                    origin = "1899-12-30")
plotmomo.atags(atags, plot.land = TRUE)
```

Based on the spatial extent of the tagging data, the next step is to define a
spatial grid. While the Kalman filter approach does not require a grid for model
fitting [@MildenbergerInPrep], using a grid enables spatial prediction,
visualization, and analysis of estimated habitat preferences and movement rates.
The `create.grid` function offers flexible functionality to define custom grids
and to manually include or exclude specific grid cells as needed.

```{r, eval=FALSE, echo=FALSE}
## Create a grid based on tagging data
grid <- create.grid(c(-150, -70), c(-30, 35),
                    dxdy = c(10,10),
                    select = 2, plot.land = TRUE)
```

Since manual selection of grid cells is not feasible in this automated vignette,
we use the grid provided in the skjepo data set, but modify it to create a
coarser resolution:

```{r, eval=TRUE, echo=FALSE}
grid <- create.grid(grid = skjepo.grid, dxdy = c(10,10))
plotmomo.grid(grid, plot.land = TRUE)
```

Another important component of *momo* is the environmental data, which
characterizes the habitat preferences of the modeled species. The `prep.env`
function formats this data into the structure required by the model:

```{r, eval=TRUE, echo=FALSE}
## Env data
env <- prep.env(skjepo.env)
plotmomo.env(env[,,1:4], plot.land = TRUE,
             xlab = "lon", ylab = "lat")
```

To speed up the fitting process in this vignette, we use a subset of the
available conventional tag data:

```{r, eval=TRUE, echo=FALSE}
ctags <- ctags[sample(1:nrow(ctags),100),]
```

The individual data sets can then be combined into a single input object using
the `setup.momo.data` function.

```{r, eval=TRUE, echo=FALSE}
## Combine and check data
dat <- setup.momo.data(grid = grid,
                       env = env,
                       ctags = ctags,
                       atags = atags)
```

The `def.conf` function generates a list of default configuration settings,
including flags that control which model functionalities are activated and how
environmental fields are aligned with the model's time steps. It is good
practice to review this configuration list to ensure that the default settings
align with the goals and structure of your analysis.

```{r, eval=TRUE, echo=FALSE}
## Default configurations
conf <- def.conf(dat)
```

Similarly, the `def.par` function generates a list of default model parameters
and their initial values. While users can adjust the initial values as needed,
caution is required when modifying parameter dimensions. These dimensions must
be consistent with the input data (e.g., the number of environmental fields) and
the configuration settings (e.g., whether passive advection, taxis, or both are
used).

```{r, eval=TRUE, echo=FALSE}
## Default parameters
par <- def.par(dat, conf)
```

With all necessary input data prepared, the model is now ready to be fitted
using *momo*.


## *Fit the model*
The model is fitted using the `fit.momo` function. Depending on the model’s
complexity and the number of tags, this step may take several minutes. To reduce
computation time, we disable the calculation of uncertainty estimates by setting
`do.sdreport = FALSE`.

```{r, eval=TRUE, echo=FALSE}
## Fitting movement model
fit <- fit.momo(dat, conf, par,
                do.sdreport = FALSE, ## only for vignette
                verbose = TRUE)
```

In addition to the input data lists, the returned object also includes the `obj`
(from RTMB) and `opt` (from the minimizer). Both contain the estimated parameter
values and other details relevant to the model fitting.

```{r, eval=TRUE, echo=FALSE}
fit$opt$par
```


## *Results*
*momo* provides several functions for visualizing model results. In this example,
since the data are simulated and the true parameters are known, we can use the
`plotmomo.compare` function to visualize both the estimated and true habitat
preferences and movement patterns.

```{r, eval=TRUE, echo=FALSE}
plotmomo.compare(sim = skjepo, fit = fit)
```


## *Advanced settings*
There are several advanced settings in *momo* that will be covered in future
vignettes. However, this vignette highlights three important ones: (i) grouping
tags into release events to speed up model fitting, (ii) using the matrix
exponential approach as an alternative to the Kalman filter, and (iii) mapping
parameters to fix or exclude them during estimation.

It is common for multiple tags to be released at the same location and time.
Grouping these tags into release events can significantly speed up the movement
modeling, as the analysis is performed per event rather than per individual tag.
The potential loss in accuracy is likely minimal and can be evaluated through
sensitivity analyses, while the grouped approach is recommended for the baseline
scenario due to its efficiency. The `get.release.event` function enables this
grouping by assigning tags to release events based on a specified spatial grid
and time vector. These events can then be added to the data list for modeling.
Note that this is step is optional.

```{r, eval=TRUE, echo=FALSE}
tmp <- get.release.events(dat,
                          grid = create.grid(dat$xrange, dat$yrange, c(1,1)),
                          time.cont = seq(dat$trange[1], dat$trange[2],
                                          1/(52*diff(dat$trange))))
dat$rel.events <- tmp$rel.events
dat$ctags$rel.event <- tmp$idx
```

By default, *momo* uses the Kalman filter approach as described in
@MildenbergerInPrep. However, it also supports an alternative method based on
the matrix exponential. You can easily switch between the two by setting the
`use.expm` flag in the configuration list. To use the matrix exponential
approach, simply set `use.expm = TRUE`.

```{r, eval=FALSE, echo=FALSE}
conf <- def.conf(dat)
conf$use.expm <- TRUE
```

Lastly, it may be useful to map parameters, for example, to estimate a single
value for both directions of passive advection. A list of default mapping
settings can be generated using the `def.map` function. The resulting map can
then be passed to `fit.momo` to control which parameters are estimated, fixed, or
shared.

```{r, eval=FALSE, echo=FALSE}
map <- def.map(dat, conf, par)
```


## *Summary*
This vignette introduced the main features and workflow of the *momo* package for
estimating animal movement and habitat preferences from tagging data. Using a
simulated data set (`skjepo`), we demonstrated how to prepare the required inputs,
configure and fit the model, and visualize the results.

We covered the preparation of conventional and archival tag data (`prep.ctags`,
`prep.atags`), environmental fields (`prep.env`), and spatial grids
(`create.grid`). These were then combined using `setup.momo.data` into a format
suitable for model fitting. Configuration and parameter initialization were
handled through `def.conf` and `def.par`, with model fitting performed using
`fit.momo`. The `plotmomo.compare` function allowed us to visualize and compare
estimated versus true movement and habitat preference patterns.

Several advanced options were introduced, including grouping tags into release
events (`get.release.event`), switching between the Kalman filter and matrix
exponential approaches (`use.expm` flag), and mapping parameters (`def.map`) to
simplify or constrain estimation.

This basic example provides a foundation for applying *momo* to real tagging
data. More details about the methodology are described in @MildenbergerInPrep.
Further details about functions and their arguments can be found in the help
files of the functions (`help()` or `?`, where the dots refer to any
function of the package). Additional vignettes will explore uncertainty
estimation, model diagnostics, and sensitivity analyses to support more robust
ecological inference and management applications.


## *References*
