---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/"
)
```

<!-- badges: start -->
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
<!-- badges: end -->

*momo* is a new R package for tagging-based *mo*vement *mo*deling,
specifically designed to estimate fine-scale animal movement patterns using
various tagging data, such as mark-recapture, mark-resight, and
data-logging/archival tags. The movement model is based on the
advection-diffusion equation and incorporates habitat preference functions,
enabling the reconstruction of individual movement paths and inference of
population-level movement dynamics using a small set of interpretable
parameters. Movement rates are estimated as a function of environmental
conditions or geographic locations.



## Installation

The current version of *momo* (v0.0.1) requires R $>= 4.0.0$ and can be
downloaded from github as follows.

::: .momo-release
```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("tokami/momo")
```
:::

The development version can be installed with:

::: .momo-devel
```{r, eval = FALSE}
remotes::install_github("tokami/momo", ref = "dev")
```
:::



## Usage

To get started with *momo*, load the package, simulate a tagging dataset, and fit
the movement model:

```{r, eval = TRUE, echo = FALSE, fig.alt = "Simulated data and model predictions"}
library(momo)

## Simulate a small tagging data set
sim <- sim.momo()

## Fit momo to the simulated data
fit <- fit.momo(sim)

## Plot simulated data and model predictions
plotmomo.compare(sim = sim, fit = fit)
```

This example illustrates the basic workflow: prepare data (here simulate), fit
the model, and access results. For real applications, momo supports multiple
types of tagging data, customizable environmental covariates, and both
estimation and prediction features.

Detailed examples and guidance are provided in the package vignettes (see
[https://tokami.github.io/momo/](https://tokami.github.io/momo/)).


## Getting help

You can find more information about *momo* on its *pkgdown* page at
[https://tokami.github.io/momo/](https://tokami.github.io/momo/). The page
includes links to vignettes, functions descriptions, version updates, and many
more. In case, your question is not answered on the *pkgdown* webpage, please
write an email to the maintainer: [Tobias
Mildenberger](mailto:t.k.mildenberger@gmail.com). In case you find bugs, please
post an issue on [here](https://github.com/tokami/momo/issues).



## Citation

Please use the R command `citation("momo")` to receive information on how to
cite this package.



## References

1. Mildenberger, T. K., Maunder, M., Nielsen, A. (in prep). momo: an R
   package for the estimation of fine-scale animal movement based on tagging
   data
2. Mildenberger, T. K., et al. (in prep). Habitat suitability and movement of
   Skipjack tuna in the Eastern Pacific Ocean
